Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-06 demystify.tex
updates

In fact, many resources are available to encourage and support students from minority and diverse groups to pursue higher education and research, e.g., specific scholarships and fellowships. Faculty in CS are also \emph{very} familiar with and are highly encouraged to support such students in their group. There are also incentives, such as dedicated funding and awards, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

Most US universities have CS studies, but many of them do not have a PhD program in CS. These universities might offer just Bachelor's degrees (e.g., BS) and no graduate studies (i.e., no MS or PhD degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has PhD in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiple locations of the University of Texas, e.g., Austin, Dallas, and Arlington, have PhDs in CS.

@@

Many students just put schools into two bins: (i) top schools that they dream about, and (ii) everything else.  They commonly use rankings from US News, which is not transparent and questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes they evaluate based on the reputation of the school's undergrad program or the reputation of the school's non-CS programs such as medical, math, or physics.
Many international students rank universities based on popular places they know in the US, e.g., California, Texas, and New York.

@@

Many students, especially international students, aim for very top schools such as Stanford and MIT, and Ivy Leagues. Every year Reddit and other forums have numerous students asking for evaluation of their chances of getting into these schools (the so called \emph{``chance me''} or \emph{``roast my CV/profile''} posts) and then later posts on being \emph{``rejected everywhere''}.

@@

Understandably, you might feel uncomfortable disclosing this, fearing you would lose a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academics and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek to work with students from diverse backgrounds and view diversity as an asset to their research group}.

@@

The DoD National Defense Science and Engineering Graduate (NDSEG) is another prestigious fellowship supporting \emph{U.S. citizens} pursuing advanced degrees in STEM fields critical to national defense (such as CS).
It provides up to three years of funding, including a  stipend, tuition coverage, and health insurance.

\subsection{Evaluation}
Unlike the NSF GRFP, for NDSEG reviewers \emph{do not} see your LoRs, i.e., we do not know who wrote them or what they wrote. Moreover, we only see \emph{redacted versions} of your submitted materials (you are asked to provide a full and redacted version of your research plan, personal essays, and CV). Thus we do not know what schools your went to, where you did your internships, your publications or research experience. We can see your current transcript, but not your past ones.

Thus, reviewers mainly get to know you and evaluate your application by reading your research plan and personal essays. The reason is for fairness and avoiding bias based on the school you went to, places you published at, your LoR writers, etc.  It could be that external reviewers like myself help with the first cut based on the redacted materials, and then the final committee makes the final decision based on the full materials.

\begin{commentbox}[Vu:]
I am used to reading full applications and thus find it a bit difficult to evaluate NDSEG applications because I do not have the full picture.  For example, a student mentioned that they have published in the topic they proposed but redacted all information about the publication, so I have no idea where they published, what the paper is about, etc.  
It is also difficult to evaluate the academic background of the student, e.g., some student just started grad school and so their transcript is essentially blank, and I do not know where they went to undergrad, etc. It is also annoying to see CVs with so many blanked out information. They might as well should just submit a blank CV one. 
\\

@@

\subsection{Common Pitfalls and Tips}
Based on my experience reviewing NDSEG applications, here are some common pitfalls to avoid:
\begin{itemize}

@@

    \item Don't make up things or exaggerate your research.  Reviewers are experts in the field and can easily spot technical inaccuracies or ideas that have already been studied. Understandbly many applicants are new, but you need to do proper research on what you propose. In particular, you should aim to have (i) clear problem statement, (ii) why it is important (esp. to DoD), (iii) what has been done and their challenges/limitations, and (iv) what you plan to do to address the challenges.  Moreover, research is often quite technical and ambitious, so you want to show some preliminary work, e.g., a small experiment by hand on some small examples, to convince reviewers that you have thought about the problem and have some initial results.\\

    Unlike GRFP where we have more things to evaluate, in NDSEG we mainly evaluate you based on your personally essays and especially research plan. So you want to make sure your research plan is well thought out and convincing.

    \item Personal stories and essays should be unique and interesting.  Many personal stories are quite generic and do not tell much about the student. For example, being an URM, involving in a robotics club or mentoring underrepresented students are quite common and do not tell much.\\

    What you want is to pick a few thing and provide concrete examples and details.  For example, provide a concrete experience in being minority and how it has affected you, talk about what you have accomplished in your robotics club, e.g., you built a robot that won a competition, or you mentored a specific URM student who then went on to do great things.  Just a small, interesting detail can make your essay stand out.

@@

\item 
personal essay / stories are written in text box and so do not llook very professional and doesn’t seem to be heavily weighted compared to research plan (e.g., very few questions asked about the personal stories compared to questions asked on research plan).

- use latex,  do not write in word and especially use left justification  (see how CS and engineer people publish papers)  
- If you have an advisor or Lor writer, ask them to look at your writing/essays .  This is even suggested directly by NSF and DoD.  You are new to the field and what your ideas might not be accurat or have already been explored — let them help you here.  While you can BS to PhD adcom because there you don’t really write a research plan,  but for these fellowships you do and reviewers are all experts in the field and would look at these carefully.

>>>>

In fact, many resources are available to encourage and support students from minority and diverse groups to pursue higher education and research, e.g., specific scholarships and fellowships. Faculty in CS are also familiar with and are encouraged to support such students in their group. There are also incentives, such as dedicated funding and awards, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

Most US universities have CS programs, but many of them do not have a CS \emph{PhD} program. These universities might offer just Bachelor's degrees (e.g., BS) and no graduate studies (i.e., no MS or PhD degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has PhD in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiple locations of the University of Texas, e.g., Austin, Dallas, and Arlington, have PhDs in CS.

@@

Many students put universities into two bins: (i) top schools that they dream about, and (ii) everything else.  They often use rankings from US News, which is not transparent and questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes they evaluate based on the reputation of the school's undergrad program or the reputation of the school's non-CS programs such as medical, math, or physics.
Many international students rank universities based on popular places they know in the US, e.g., California, Texas, and New York.

@@

Many international students aim for very top schools such as Stanford and MIT, and Ivy Leagues. Every year Reddit and other forums have numerous students asking for evaluation of their chances of getting into these schools (the so called \emph{``chance me''} or \emph{``roast my CV/profile''} posts) and then later posts on being \emph{``ghosted and rejected everywhere''}.

@@

Understandably, you might feel uncomfortable disclosing this, fearing you would lose a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academics and likely would not be tolerated by the university). In my experience, it is often the opposite: \emph{many profs. actively seek to work with students from diverse backgrounds and view diversity as an asset to their research group}.

@@

The DoD National Defense Science and Engineering Graduate (NDSEG) is another prestigious fellowship supporting \emph{U.S. citizens} pursuing advanced degrees in STEM fields \emph{critical to national defense} (such as CS).
It provides up to three years of funding, including a  stipend, tuition coverage, and health insurance.

You can read more details about NDSEG and its guideline for applicants from the \href{https://www.ndsegfellowships.org/}{official website}. Here, I mainly focus on things that stand out to me as reviewers and also compare them with NSF GRFP as many students apply to both.
\subsection{Evaluation}

\paragraph{Redacted info.} Unlike the NSF GRFP, for NDSEG reviewers \emph{do not} see your LoRs, i.e., we do not know who wrote them or what they wrote. Moreover, we only see \emph{redacted versions} of your submitted materials (applicants are asked to provide full and redacted versions of your research plan, CV, and other materials). Thus we do not know what schools your went to, where you did your internships, your publications or research experience.

Thus, reviewers mainly get to know you and evaluate your application by reading your research plan and personal essays. This is for \emph{fairness and avoiding bias} based on the school you went to, places you published at, your LoR writers, etc.  Nonethless, I felt I am missing too much information. However, it could be that external reviewers like myself help with the first cut based on the redacted materials, and then there are internal reviewers who makes the final decision based on the full materials.

\paragraph{Non-expert} NDSEG reviewers hold advanced degrees in STEM fields. However, NDSEG reviewers are more general and might not be in your specific field. For example, in GRFP I mostly review applications in PL and FM, while in NDSEG it seems that I review applications in all areas of CS and even non-CS.  Thus, you should make your research plan \emph{more accessible to a general audience}.

\begin{commentbox}[Vu:]
I find a bit difficult to evaluate NDSEG applications because I do not have the full picture.  For example, a student mentioned that they have published in the topic they proposed but redacted all information about the publication, so I have no idea where they published, what the paper is about, etc.  
It is also difficult to evaluate the academic background of the student, e.g., some student just started grad school and so their transcript is essentially blank, and I do not know where they went to undergrad, etc. It is also annoying to see CVs with so many blanked out information. They might as well should just submit a blank CV one. 
\\

@@

\subsection{Common Pitfalls and Tips}
Based on my experience in reviewing, here are some common pitfalls to avoid:
\begin{itemize}

@@

    \item Unclear research plan. 
     You should aim to have (i) clear problem statement, (ii) why it is important (esp. to DoD), (iii) what has been done and their challenges/limitations, and (iv) what you plan to do to address the challenges.  Moreover, research is often quite technical and ambitious, so you want to show some preliminary work, e.g., a small experiment by hand on some small examples, to convince reviewers that you have thought about the problem and have some initial results.\\

    Unlike GRFP where we have more things to evaluate, in NDSEG we mainly evaluate you based on your personally essays and especially research plan. So you should do proper research on what you propose and make it clear and convincing.

    \item Personal stories and essays should be unique and interesting. Many essays and stories I read are quite generic and do not tell much about the student. For example, being an URM, involving in a robotics club or mentoring underrepresented students are quite common and do not tell much.\\

    You should pick a few thing and provide concrete examples and details.  For example, provide a concrete experience in being minority and how it has affected you, talk about what you have accomplished in your robotics club, e.g., you built a robot that won a competition, or you mentored a specific URM student who then went on to do great things.  Just a small, interesting detail can make your essay stand out.

@@

%personal essay / stories are written in text box and so do not llook very professional and doesn’t seem to be heavily weighted compared to research plan (e.g., very few questions asked about the personal stories compared to questions asked on research plan).

%- use latex,  do not write in word and especially use left justification  (see how CS and engineer people publish papers)  
\item If you have an advisor or Lor writer, ask them to look at your writing/essays. This is because you are new to the field and research writing, and your ideas might not be accurate or have already been explored.
